A Short Guide to Anyons and Modular Functors

### 2016-09-24 guide.tex
commit edits

\documentclass[aps, letterpaper, onecolumn, superscriptaddress, notitlepage, 10pt]{revtex4-1}
\usepackage{times}

@@

\title{A Short Guide To Modular Functors, Anyons, etc.}

\author{Simon Burton}
\affiliation{Centre for Engineered Quantum Systems, School of Physics, 
The University of Sydney, Sydney, Australia}
%\author{Courtney G.\ Brell}
%\affiliation{Institut f\"{u}r Theoretische Physik, Leibniz Universit\"{a}t Hannover, 
%Appelstra\ss{}e 2, 30167 Hannover, Germany}
%%\affiliation{Perimeter Institute for Theoretical Physics, Waterloo, Ontario, Canada}
%\author{Steven T.\ Flammia}
%\affiliation{Centre for Engineered Quantum Systems, School of Physics, 
%The University of Sydney, Sydney, Australia}

@@

@@

In this section we will describe abstractly the theory of two-dimensional topologically ordered systems
with anyonic excitations.
There are two main approaches to defining these,
one being more algebraic and the other more topological.

The algebraic side is known to mathematicians as the study of braided
tensor categories, or more specifically, modular tensor categories.
This algebraic language appears to be more commonly used 
in the physics literature~\cite{Kitaev2006}.
Such algebraic calculations can be interpreted as
manipulations of string diagrams~\cite{Baez2010},
or \emph{skeins}, which begins to hint at
how knot theory and topology comes into play.

Coming from the other direction, one starts with
a topological space (of low dimension) and attempts
to extract a combinatorial or algebraic description of how this
space can be built from joining smaller (simpler) pieces together.
%This is perhaps more difficult to grapple with.
These topologically rooted constructs are known as modular functors, 
or the closely related topological quantum field theories (TQFT's).
%The much older study of Homology also extracts algebraic
%information out of topological spaces, but

It is not obvious a priori that these two approaches -- algebraic and topological -- should meet, hinting at a deep underlying structure.
%That these two approaches -- algebraic versus topological -- meet is one of the great
%miracles of modern mathematics and physics.

@@

In this section we try to motivate 
the formalism of modular functor presented
below.
Our focus will be to show some of the close connections between
the algebraic and topological approaches to
particle exchange statistics.

In three spatial dimensions, the process of winding one
particle around another, a \emph{monodromy}, is topologically trivial.
This is because the path can be deformed back to the
identity; there is no obstruction:
\begin{center}
\includegraphics[]{pic-monodromy3d.pdf}
\end{center}
The square root of this operation is a swap:
\begin{center}
\includegraphics[]{pic-swap.pdf}
\end{center}
For identical particles this is a symmetry of the system. 
%%%%%% \cggb{What does symmetry mean here? Since a swap of non-bosons will pick up a phase.}
Continuing with this line of thought 
leads to consideration of the \emph{symmetric group}
on $n$ letters, $S_n$. 
This group is generated by the $n-1$ swap operations
$s_1,...,s_{n-1}$ that obey the relations
\begin{align*}
    s_i^2 &= 1, \\
    s_i s_j &= s_j s_i \ \ \ \mbox{for}\ \ |i-j|>1,\\
    s_i s_{i+1} s_i &= s_{i+1} s_i s_{i+1} \ \ \ \mbox{for}\ \ 1\le i \le n-2.
\end{align*}
This group acts on the statespace of $n$ particles $\F_n$
via unitaries $U(\F_n)$:
$$
    \Phi : S_n \to U(\F_n).
$$

When the above monodromy is constrained to two dimensions
we can no longer deform this process to the identity:
\begin{center}
\includegraphics[]{pic-monodromy2d.pdf}
\end{center}
and so in two dimensions we cannot expect a swap to square to the identity.
To see this more clearly,
we consider the world-lines of these particles.
Here we show three particles undergoing an exchange and then
returning to the original location:
\begin{center}
\includegraphics[]{pic-braid-worldlines.pdf}
\end{center}
Note that if we allow the particles to move in one extra
dimension then we can untangle these braided world lines.
%The question is, how has the wavefunction of the system
%changed from $t_0$ to $t_1$?
The question is now, what is the group that acts on the
state space from $t_0$ to $t_1$?
Examining the structure of these processes more closely,
we see that we can compose them by sequentially performing
two such braids, one after the other:
\begin{center}
\includegraphics[]{pic-braid-compose.pdf}
\end{center}
And, by ``reversing time'', we can undo the effect of
any braid:
\begin{center}
\includegraphics[]{pic-braid-inverse.pdf}
\end{center}
This shows that these processes do form a group,
known as the \emph{braid group.}
For $n$ particle worldlines we denote this group as $B_n.$
For identicle particles this group acts as symmetries of the
state space:
%Now we would expect to see an action on the statespace $\F_n:$
$$
    \Phi : B_n \to U(\F_n).
$$

@@

%The easiest way to understand these relations is to consider the
It is easy to see that the geometric braid group satisfies these
relations. Here we show the braids corresponing to the generators $\sigma_i:$
\begin{center}
\includegraphics[]{pic-braid-sigma.pdf}
\end{center}
So that 
$\sigma_i \sigma_j = \sigma_j \sigma_i \ \ \ \mbox{for}\ \ |i-j|>1$
because the two braids are operating on disjoint strands.
The second relation is also easy to see:
\begin{center}
\includegraphics[]{pic-braid-YB.pdf}
\end{center}

There is another important geometric representation of
the braid group which is purely two-dimensional.
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms 
%\cggb{why diffeomorphisms and not homeomorphisms?}
$f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
that measure the charge on the
interior of the curve.
The importance of understanding the
braid group as identicle to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivelant, but the actual proof of this is somewhat involved.
We refer the reader to \cite{Kassel2010} for an excellent contemporary
reference that fills in these gaps.

@@

So far we have been studying the exchange statistics
for $n$ identical particles. Without this restriction, one needs to
constrain the allowed exchange processes so as to preserve particle type.
For example, if all particles are different we would use the
\emph{pure braid group} $PB_n.$
The geometric description of
this group is as the fundamental group of the labelled
configuration space
$$
    PB_n = \pi_1 ( \Conf_n, x ).
$$
Loops in $\Conf_n$ correspond to braids where each strand returns
to the point it started from.
The algebraic description of this group is somewhat complicated and
we omit this.
In terms of the mapping class group, we can describe $PB_n$
as the \emph{pure mapping class group}:
$$
    PMCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(x)=x \ \ \mbox{for}\ x\in Q_n\} / \sim_{\mbox{iso}}.
$$
%we can write generators of this group in terms of Dehn twists which
%we discuss below.

@@

Here we show four 
geometric approaches to representing a twist.
Anyone that has struggled to untangle their headphone cable
will immediately see what is going on here.
On the left we have a loop; it is not a braid because it
travels backwards in time.
If we pull on this loop to make it straight, we introduce a twist.
This is shown in the next figure, 
where we show a \emph{framing} which is a non-degenerate 
vector field along the world-lines of a braid.
The initial and final vectors in the vector field must be the same.
By non-degenerate we mean that the vector field is everywhere
non-zero and also not tangent to the world-line.
In the next figure we show a \emph{ribbon}:
instead of point particles we have short one-dimensional curves in $M$.
%This is another way to encode a twist without using the differential
%structure of $M.$
On the right the particle is represented as a boundary
component of $M$ with a distinguished point.
In this picture the world-line looks like a tube.
%It is not difficult to see that these 
All these representations of twists carry
essentially the same information.
\begin{center}
\includegraphics[]{pic-framed.pdf}
\end{center}
In the sequal we will stick
to thinking of particles as boundary
components because this fits well
with the way we are formulating 
observables as simple closed curves in $M$.
%More concretely,
%we augment the configuration space
%with tangent vectors. For each point $x_i$
%in a configuration $(x_1,...,x_n)\in \prod M$ we
%attach a non-zero tangent vector $v_i\in T_{x_i}M.$
%In terms of a mapping class group, we will
%consider not just a manifold $M$ with
%distinguished points $Q_n,$ but a manifold with $n$ disjoint
%boundary componends, or ``holes''.
%Each of these boundary components comes
%with a distinguished point.

>>>>

\documentclass[aps, letterpaper, onecolumn, superscriptaddress, notitlepage, 11pt, groupedaddress]{revtex4-1}
\usepackage{times}

@@

\title{A Short Guide to Anyons and Modular Functors}

\author{Simon Burton}
\affiliation{Centre for Engineered Quantum Systems, School of Physics, 
The University of Sydney, Sydney, Australia}

@@

\begin{abstract}
To the working physicist, anyon theory is meant to describe
certain quasi-particle excitations occuring in two dimensional
topologically ordered systems. 
A typical calculation using this theory will involve operations
such as $\otimes$ to combine anyons, 
$F^{abc}_{d}$ to re-associate such combinations,
and $R^{ab}_c$ to commute or braid these anyons. 
Although there is a powerful string-diagram notation that greatly assists these
manipulations, we still appear to be operating on particles
arranged on a one-dimensional line, algebraically ordered from left to right.
The obvious question is, where is the other dimension?
The topological framework for considering these anyons as truly living in a
two dimensional space is known as a modular functor, or topological
quantum field theory. 
In this work we show how the apparently one-dimensional algebraic anyon theory
is secretly the theory of anyons living in a fully two-dimensional system.
The mathematical literature covering this secret is vast, 
and we try to distil this down into something more manageable.
%While avoiding much of the
%category theoretic jargon found in the mathematical literature.
%To the physicist, anyons are manipulated algebraically, using operations
%such as $\otimes$ to combine, $F^{abc}_{de}$ to re-associate, and $R^{ab}_c$
%to commute or braid. 
%Although there are string-diagram notations that greatly assist these
%manipulations, such calculations still appear to be operating on particles
%arranged in a one-dimensional line, algebraically ordered from left to right.
%The correct notion for considering these anyons as truly living in a
%two dimensional space is known as a modular functor, or topological
%quantum field theory. 
%In this work we show how the apparently one-dimensional algebraic theory
%is secretly a fully-two dimensional theory, while avoiding much of the
%category theoretic jargon found in the mathematical literature.
\end{abstract}

@@

In this work we describe the theory of two-dimensional topologically ordered systems
with anyonic excitations. 
There are two main approaches to defining these,
one being more algebraic and the other more topological.

The algebraic side is known to mathematicians as the study of braided
tensor categories, or more specifically, modular tensor categories.
This algebraic language appears to be more commonly used 
in the physics literature, such as the well cited Appendix E of Kitaev~\cite{Kitaev2006}.
Such algebraic calculations can be interpreted as
manipulations of string diagrams~\cite{Baez2010},
or \emph{skeins}.
These strings encode connections in the algebra, for example 
the Einstein summation convention. But they also faithfully
encode the twisting that occurs when anyons are re-ordered or
braided around each other. And these kinds of spatial relationships
are fruitfully studied using topological methods.

Working from the other direction, one starts with
a topological space (of low dimension) and attempts
to extract a combinatorial or algebraic description of how this
space can be built from joining smaller (simpler) pieces together.
%This is perhaps more difficult to grapple with.
These topologically rooted constructs are known as modular functors, 
or the closely related topological quantum field theories (TQFT's).
%The much older study of Homology also extracts algebraic
%information out of topological spaces, but

%It is not obvious a priori that these two approaches -- algebraic and topological -- should meet, hinting at a deep underlying structure.
That these two approaches -- algebraic versus topological -- meet is one of the great
surprises of modern mathematics and physics.

@@

In this section we begin with the familiar question 
of particle exchange in three dimensions, whose answer is bosons and fermions.
We then show how in restricting the particles to two dimensions many
more possibilities arise.
Our focus will be on the close connection between the algebraic and the topological viewpoints,
aiming to motivate the definition of a modular functor given in the next section.

In three spatial dimensions, the process of winding one
particle around another, a \emph{monodromy}, is topologically trivial.
This is because the path can be deformed back to the
identity; there is no obstruction:
\begin{center}
\includegraphics[]{pic-monodromy3d.pdf}
\end{center}
The square root of this operation is a swap:
\begin{center}
\includegraphics[]{pic-swap.pdf}
\end{center}
For identical particles this is a symmetry of the system. 
%%%%%% \cggb{What does symmetry mean here? Since a swap of non-bosons will pick up a phase.}
Continuing with this line of thought 
leads to consideration of the \emph{symmetric group}
on $n$ letters, $S_n$. 
This group is generated by the $n-1$ swap operations
$s_1,...,s_{n-1}$ that obey the relations
\begin{align*}
    s_i^2 &= 1, \\
    s_i s_j &= s_j s_i \ \ \ \mbox{for}\ \ |i-j|>1,\\
    s_i s_{i+1} s_i &= s_{i+1} s_i s_{i+1} \ \ \ \mbox{for}\ \ 1\le i \le n-2.
\end{align*}
%This group acts on the statespace of $n$ particles $\F_n$
%via unitaries $U(\F_n)$:
%$$
%    \Phi : S_n \to U(\F_n).
%$$
Writing the Hilbert space of the system as $V$ 
we would then expect $S_n$ to act on this space via unitary transformations $U(V):$
$$
    \F : S_n \to U(V).
$$
% For bosons this is ... for fermions it is ...
This is the first example of the kind of functor we will be talking about. 
In this case it is a group representation; $\F$ is a homomorphism between
two groups.

When the above monodromy is constrained to two dimensions
we can no longer deform this process to the identity:
\begin{center}
\includegraphics[]{pic-monodromy2d.pdf}
\end{center}
and so in two dimensions we cannot expect a swap to square to the identity.
To see this more clearly,
we must examine the entire (2+1)-dimensional \emph{world-lines} of these particles.
For an example, here we show the world-lines of three particles undergoing an exchange and then
returning to their original positions:
\begin{center}
\includegraphics[]{pic-braid-worldlines.pdf}
\end{center}
Note that if we allow the particles to move in one extra
dimension then we can untangle these braided world lines.
%The question is, how has the wavefunction of the system
%changed from $t_0$ to $t_1$?
The question is now, what is the group that acts on the
state space from $t_0$ to $t_1$?
Examining the structure of these processes more closely,
we see that we can compose them by sequentially performing
two such braids, one after the other:
\begin{center}
\includegraphics[]{pic-braid-compose.pdf}
\end{center}
And, by ``reversing time'', we can undo the effect of
any braid:
\begin{center}
\includegraphics[]{pic-braid-inverse.pdf}
\end{center}
This shows that these processes do form a group,
known as the \emph{braid group.}
For $n$ particle worldlines we denote this group as $B_n.$
For identicle particles this group acts as symmetries of the
state space:
%Now we would expect to see an action on the statespace $\F_n:$
$$
%    \Phi : B_n \to U(\F_n).
    \F : B_n \to U(V).
$$

@@

%The easiest way to understand these relations is to consider the
It is easy to see that the geometric braid group satisfies these
relations. Here we show the braids corresponing to the generators $\sigma_i:$
\begin{center}
\includegraphics[]{pic-braid-sigma.pdf}
\end{center}
So that 
$\sigma_i \sigma_j = \sigma_j \sigma_i \ \ \ \mbox{for}\ \ |i-j|>1$
because the two braids are operating on disjoint world-lines.
The second relation is also easy to see:
\begin{center}
\includegraphics[]{pic-braid-YB.pdf}
\end{center}

There is another important geometric representation of
the braid group which is purely two-dimensional.
We pick an $n$-point finite subset $Q_n \subset M$
and consider diffeomorphisms 
%\cggb{why diffeomorphisms and not homeomorphisms?}
$f : M \to M$
that map $Q_n$ to itself.
The \emph{mapping class group} of $M$ relative to $Q_n$
is the set of all such diffeomorphisms up to an equivalence relation $\sim_{\mbox{iso}}$:
$$
    MCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(Q_n)=Q_n \} / \sim_{\mbox{iso}}
$$
The equivalence relation $\sim_{\mbox{iso}}$ is called \emph{isotopy}
which allows for any continuous deformation of $f:M\to M$ that
fixes each point in $Q_n.$
Each generator $\sigma_i$ of the braid group is found in $MCG(M, Q_n)$
as a \emph{half-twist} that swaps two points $i, i+1 \in Q_n$:
\begin{center}
\includegraphics[]{pic-halftwist.pdf}
\end{center}
In order to show
the action of
this half-twist we have 
decorated the manifold with
a a checkered pattern,
but there is a more important
object that lives on the manifold
itself.
An \emph{observable} is a simple
closed curve in $M$ that does
not intersect $Q_n.$
Such a closed curve is called an
observable because these will be
associated to measurements
that measure the charge on the
interior of the curve.
The importance of understanding the
braid group as identicle to the mapping class group
is now manifest:
whereas geometric braids act on states
as in a Schr\"{o}dinger picture,
elements of the mapping class group
act on the observables as in
a Heisenberg picture.
\begin{center}
\includegraphics[]{pic-heisenberg.pdf}
\end{center}
This diagram should give the reader some idea as
to why these two definitions of the braid group are
equivelant, but the actual proof of this is somewhat involved.
We refer to \cite{Kassel2010} for an excellent contemporary
reference that fills in these gaps.

@@

So far we have been studying the exchange statistics
for $n$ identical particles. Without this restriction, one needs to
constrain the allowed exchange processes so as to preserve particle type.
For example, if all particles are different we would use the
\emph{pure braid group} $PB_n.$
The geometric description of
this group is as the fundamental group of the labelled
configuration space
$$
    PB_n = \pi_1 ( \Conf_n, x ).
$$
Loops in $\Conf_n$ correspond to braids where each world-line returns
to the point it started from.
The algebraic description of this group is somewhat complicated and
we omit this.
In terms of the mapping class group, we can describe $PB_n$
as the \emph{pure mapping class group}:
$$
    PMCG(M, Q_n) = \{ f : M \to M \ \ \mbox{such that}\ \ f(x)=x \ \ \mbox{for}\ x\in Q_n\} / \sim_{\mbox{iso}}.
$$
%we can write generators of this group in terms of Dehn twists which
%we discuss below.

@@

Here we show four 
geometric approaches to representing a twist.
Any person that has struggled to untangle their headphone cable
will immediately see what is going on here.
\begin{center}
\includegraphics[]{pic-framed.pdf}
\end{center}
On the left we have a loop; it is not a braid because it
travels backwards in time.
If we pull on this loop to make it straight, we introduce a twist.
This is shown in the next figure, 
where we show a \emph{framing} which is a non-degenerate 
vector field along the world-lines of a braid.
The initial and final vectors in the vector field must be the same.
By non-degenerate we mean that the vector field is everywhere
non-zero and non-tangent to the world-line.
In the next figure we show a \emph{ribbon}:
instead of point particles we have short one-dimensional curves in $M$.
%This is another way to encode a twist without using the differential
%structure of $M.$
On the right the particle is represented as a boundary
component (a hole) of $M$ with a distinguished point.
In this picture the world-line looks like a tube.
%It is not difficult to see that these

All these representations of twists carry
essentially the same information.
In the sequal we will stick
to thinking of particles as boundary
components because this fits well
with the way we are formulating 
observables as simple closed curves in $M$.
%More concretely,
%we augment the configuration space
%with tangent vectors. For each point $x_i$
%in a configuration $(x_1,...,x_n)\in \prod M$ we
%attach a non-zero tangent vector $v_i\in T_{x_i}M.$
%In terms of a mapping class group, we will
%consider not just a manifold $M$ with
%distinguished points $Q_n,$ but a manifold with $n$ disjoint
%boundary componends, or ``holes''.
%Each of these boundary components comes
%with a distinguished point.
